\documentclass{article}
\usepackage[preprint]{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{enumitem}
\usepackage[most]{tcolorbox}
\usepackage{natbib}

\title{Beyond Aggregate Scores: Diagnosing the Explicit-to-Applied\\Theory of Mind Gap in Large Language Models}

\author{
  Anonymous Author(s)
}

\begin{document}
\maketitle

\begin{abstract}
Aggregate Theory-of-Mind (ToM) scores in large language models are hard to interpret because they often mix two different abilities: \emph{recognizing} an agent's mental state and \emph{applying} that state to predict behavior or evaluate actions.
We formalize this explicit-to-applied gap, operationalize it across four ToM benchmarks, and introduce Belief-State Test-Time Training (BSTTT), a diagnostic protocol for distinguishing latent parameter-level competence from inference-time failure.
BSTTT adapts lightweight LoRA modules on answered support questions from each test episode and compares a ToM-aligned action-reconstruction objective against a matched generic next-token objective.
Across preserved multi-seed evaluations on Qwen-2.5-7B, Llama-3.1-8B, and GPT-oss-20B, three patterns recur:
\textbf{(i)}~recovery is domain-specific---action-reconstruction adaptation improves applied performance by up to 15 percentage points, whereas matched next-token adaptation yields little or no benefit;
\textbf{(ii)}~prompting and weight adaptation are complementary, with composed interventions outperforming either class alone when the benchmark exposes a recoverable gap;
\textbf{(iii)}~DynToM-style temporal belief tracking remains resistant to all tested interventions, indicating a distinct bottleneck rather than a simple application failure.
The main implication is evaluative: aggregate ToM scores can overstate applied social reasoning while simultaneously hiding recoverable latent competence.
We therefore recommend decomposed reporting, explicit temporal-reasoning axes, and adaptation-based diagnostics for LLM social-reasoning benchmarks.
\end{abstract}


%% ===== SECTION 1: INTRODUCTION =====
\section{Introduction}
\label{sec:intro}

What conclusion should we draw from a high Theory-of-Mind (ToM) benchmark score?
A model may correctly state that Mary falsely believes the yogurt in her fridge is still fresh, yet fail to predict that she will eat it---or fail to judge that doing so was rational from her perspective~\citep{gu2024simpletom}.
The evaluative failure is not simply missing mental-state recognition; it is the failure to \emph{apply} a recognized mental state to behavior and judgment.

Current ToM benchmarks often obscure this distinction.
Many report a single aggregate accuracy that combines mental-state identification questions (\emph{``What does Mary believe?''}) with behavioral or evaluative questions (\emph{``What will Mary do?''})~\citep{kosinski2024evaluating, strachan2024testing, sap2022neural}.
When a model scores 80\% on such a benchmark, we cannot tell whether it is broadly competent or whether strong explicit recognition is being averaged with much weaker application.
Across the benchmarks we study, we repeatedly observe large explicit--applied gaps under standard inference.
As a result, aggregate scores can overstate the downstream-relevant social reasoning that matters for anticipating actions, consequences, and belief-conditioned rationality.

This explicit-to-applied gap leads to a more specific diagnostic question: does the gap reflect a \emph{capacity limitation}---the model lacks the internal machinery needed to bridge from recognition to prediction---or a \emph{latent competence} that standard autoregressive decoding fails to activate?
The distinction matters because it changes the research prescription.
Capacity limits point toward additional training or architectural change; latent competence points toward better inference-time strategies and better evaluation practice.
Prompt-level interventions---chain-of-thought~\citep{wei2022chain}, perspective-taking~\citep{wilf2024think}, symbolic belief tracking~\citep{sclar-etal-2023-minding}---can improve performance, but they do not resolve this question because they modify the model's input rather than testing whether the relevant knowledge is already encoded in the parameters~\citep{gu2024simpletom}.

To answer that question, we introduce a diagnostic evaluation protocol that intervenes directly on model parameters.
Our protocol, \textbf{Belief-State Test-Time Training (BSTTT)}, adapts lightweight LoRA modules~\citep{hu2022lora} on answered support questions drawn from each test episode and then evaluates held-out applied ToM questions from the same scenario.
The key methodological idea is a \emph{contrastive control}: we compare a ToM-aligned objective (action reconstruction, which trains the model to predict behaviorally correct answers) against a generic objective (next-token prediction on the same text) under identical adaptation conditions---same model, same support data, same parameters, same gradient budget.
If recovery appears only under the aligned objective, the evidence favors a domain-specific belief-to-behavior mapping already latent in the parameters rather than generic exposure to the episode.
This turns test-time training from a performance intervention into a \emph{diagnostic instrument} for interpreting benchmark failures.

We evaluate three open-weight instruction-tuned models (Qwen-2.5-7B, Llama-3.1-8B, GPT-oss-20B) on four ToM benchmarks spanning static and temporal scenarios, first-order and higher-order beliefs, and recognition and application (SimpleToM, Hi-ToM, OpenToM, DynToM).
We compare frozen baselines, structured prompting, BSTTT with both objectives, and composed variants that pair weight adaptation with prompting.
Three findings emerge:
%
\begin{enumerate}[leftmargin=*,itemsep=2pt,topsep=3pt]
    \item \textbf{The gap reflects latent competence, not absent knowledge.}
    Action-reconstruction adaptation closes applied-performance gaps by up to 15 percentage points on Hi-ToM and lifts behavior prediction from 0\% to 54\% on SimpleToM.
    Matched next-token adaptation provides little or no benefit.

    \item \textbf{Weight-level and input-level interventions are complementary.}
    Combining BSTTT with structured prompting consistently outperforms either component alone when the benchmark exposes recoverable latent competence.

    \item \textbf{Temporal belief tracking is a genuine capacity bottleneck.}
    On DynToM, which requires tracking how beliefs evolve across multi-event trajectories, all methods---including oracle-level structured support---yield only modest improvements ($\leq$3.8~pp).
\end{enumerate}

Our contributions are as follows:
%
\begin{enumerate}[leftmargin=*,itemsep=2pt,topsep=3pt]
    \item \textbf{Evaluation framework.}
    We formalize the explicit-to-applied ToM distinction (Eq.~\ref{eq:gap}) and operationalize it across four structurally diverse benchmarks, providing a reusable decomposition that reveals systematic gaps hidden by aggregate scoring.

    \item \textbf{Diagnostic protocol.}
    We introduce BSTTT as a controlled diagnostic probe that distinguishes capacity limitations from inference-time failures through a contrastive comparison of domain-aligned and domain-generic self-supervision.

    \item \textbf{Cross-benchmark empirical analysis.}
    Across three model families and four benchmarks, we establish the domain-specificity, composability, and structural-variation findings, showing how different evaluation setups support different conclusions about model competence.

    \item \textbf{Actionable recommendations.}
    We provide concrete guidelines for future ToM evaluation: separately report explicit and applied accuracy, include temporal reasoning as a distinct evaluation axis, and use adaptation-based diagnostics to estimate the upper bound of model competence.
\end{enumerate}


%% ===== SECTION 2: THE GAP =====
\section{The Explicit-to-Applied Theory of Mind Gap}
\label{sec:gap}

We begin by formalizing the distinction that motivates our evaluation.
Consider a ToM benchmark $\mathcal{B}$ consisting of question--answer pairs $(q_i, a_i)$ grounded in narrative social scenarios.
We partition $\mathcal{B}$ into two subsets based on what cognitive operation each question requires:
%
\begin{itemize}[leftmargin=*,itemsep=1pt,topsep=3pt]
    \item \textbf{Explicit questions} $\mathcal{B}_\mathrm{E}$: \emph{Can the model identify an agent's mental state?}
    These probe beliefs, emotions, intentions, or awareness that are stated or straightforwardly inferable from the narrative.
    \item \textbf{Applied questions} $\mathcal{B}_\mathrm{A}$: \emph{Can the model use that mental state to predict behavior, anticipate consequences, or evaluate actions?}
    These require integrating the recognized mental state with causal and commonsense world knowledge.
\end{itemize}
%
Let $\mathrm{acc}(\mathcal{M}, \mathcal{B}_\mathrm{E})$ and $\mathrm{acc}(\mathcal{M}, \mathcal{B}_\mathrm{A})$ denote a model $\mathcal{M}$'s accuracy on each subset.
The \textbf{explicit-to-applied gap} is:
%
\begin{equation}
    \Delta_\mathrm{gap}(\mathcal{M}, \mathcal{B}) \;=\; \mathrm{acc}(\mathcal{M}, \mathcal{B}_\mathrm{E}) \;-\; \mathrm{acc}(\mathcal{M}, \mathcal{B}_\mathrm{A})
    \label{eq:gap}
\end{equation}
%
A positive $\Delta_\mathrm{gap}$ indicates that the model recognizes mental states more reliably than it applies them; $\Delta_\mathrm{gap} \approx 0$ indicates balanced competence.

\subsection{Operationalization Across Benchmarks}
\label{sec:gap_operationalization}

We operationalize this decomposition across four benchmarks that span the major axes along which ToM has been evaluated (Table~\ref{tab:gap_operationalization}).

\begin{table}[t]
    \caption{Operationalization of the explicit--applied decomposition. Each benchmark's questions are partitioned into explicit (mental-state recognition) and applied (behavioral inference) subsets. Hi-ToM contains exclusively applied questions; we use BSTTT adaptation recovery as an indirect gap measure.}
    \label{tab:gap_operationalization}
    \centering
    \small
    \begin{tabular}{@{}lllll@{}}
        \toprule
        \textbf{Benchmark} & \textbf{Explicit ($\mathcal{B}_\mathrm{E}$)} & \textbf{Applied ($\mathcal{B}_\mathrm{A}$)} & \textbf{Format} & \textbf{$|\mathcal{B}|$} \\
        \midrule
        SimpleToM   & Mental-state QA       & Behavior QA, Judgment QA              & 2-way MC & 60 \\
        DynToM      & Understanding (type\_a) & Transformation (type\_c, type\_d)     & MC (7--8) & 200 \\
        Hi-ToM      & \emph{(indirect)}     & Recursive belief, location             & 15-way MC & 1{,}200 \\
        OpenToM     & Attitude QA           & Multihop first-order QA                & 3/6-way MC & 2{,}000 \\
        \bottomrule
    \end{tabular}
\end{table}

\textbf{SimpleToM}~\citep{gu2024simpletom} provides the cleanest operationalization: each scenario generates a matched triplet---mental-state identification (explicit), behavior prediction (applied), and judgment (applied)---grounded in the same narrative.
\textbf{DynToM}~\citep{xiao-etal-2025-towards} extends the decomposition temporally: understanding questions probe current mental state, while transformation questions require tracking belief evolution across five events.
\textbf{Hi-ToM}~\citep{wu-etal-2023-hi} tests higher-order recursive reasoning with all questions applied; we characterize the gap indirectly via adaptation recovery.
\textbf{OpenToM}~\citep{xu-etal-2024-opentom} separates attitude inference from multihop first-order reasoning requiring multi-step chaining.
This decomposition is operational and benchmark-specific: it is intended as an evaluation device for separating recognition from application, not as a universal cognitive taxonomy of ToM.


\subsection{Diagnostic Question}
\label{sec:diagnostic_question}

A positive $\Delta_\mathrm{gap}$ under frozen inference admits two competing explanations:
%
\begin{description}[leftmargin=1.5em,labelwidth=1.5em,itemsep=2pt]
    \item[\textbf{H1} (Capacity deficit):] The model lacks the internal representations needed to bridge from recognition to prediction. The gap is irreducible without additional training or architectural change.
    \item[\textbf{H2} (Latent competence):] The model's parameters encode sufficient social knowledge, but standard decoding fails to activate the belief-to-behavior mapping. The gap reflects an inference-time failure.
\end{description}
%
Prompt-level interventions cannot resolve this distinction: they modify what the model \emph{sees} but do not reveal whether the model's \emph{parameters} encode the requisite knowledge.
Resolving H1 versus H2 requires a controlled intervention on model parameters---one targeted enough to attribute any recovery to a specific causal mechanism.


%% ===== SECTION 3: DIAGNOSTIC PROTOCOL =====
\section{Diagnostic Protocol: Belief-State Test-Time Training}
\label{sec:bsttt}

The diagnostic question requires an intervention that modifies model parameters in a controlled manner and measures whether applied ToM accuracy recovers.
We introduce Belief-State Test-Time Training (BSTTT), an episodic adaptation protocol designed specifically for this purpose.

\subsection{Episode Design}
\label{sec:episode_design}

BSTTT operates episodically.
Each \textbf{episode} $e = (\mathcal{S}, q)$ consists of a \textbf{support set} $\mathcal{S} = \{(x_j, a_j^*)\}_{j=1}^{|\mathcal{S}|}$ of answered questions and a held-out \textbf{query} $q = (x_q, a_q^*)$, both drawn from the same social scenario.
No data external to the benchmark is used.

Episode construction is benchmark-specific (Table~\ref{tab:episode_construction}):

\begin{table}[t]
    \caption{Episode construction across benchmarks.}
    \label{tab:episode_construction}
    \centering
    \small
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Benchmark} & \textbf{Support source} & \textbf{Query target} & \textbf{$|\mathcal{S}|$} \\
        \midrule
        SimpleToM & Parallel Qs from same scenario & Behavior / Judgment & 1 \\
        DynToM    & Preceding Qs in temporal sequence & Transformation & $\leq$4 \\
        Hi-ToM    & Other Qs from same narrative & Recursive belief & $\leq$4 \\
        OpenToM   & Other Qs from same narrative & Multihop first-order & $\leq$4 \\
        \bottomrule
    \end{tabular}
\end{table}


\subsection{Experimental Conditions: Two Adaptation Objectives}
\label{sec:objectives}

The diagnostic power of BSTTT lies in comparing two adaptation objectives under identical conditions---same model, same LoRA parameters, same support data, same gradient steps.

\paragraph{Action Reconstruction (AR).}
The ToM-aligned condition trains the model to predict the behaviorally correct answer:
\begin{equation}
    \mathcal{L}_\mathrm{AR} = -\frac{1}{|\mathcal{S}|} \sum_{j=1}^{|\mathcal{S}|} \log \frac{\exp(s_{a_j^*})}{\sum_{c=1}^{C} \exp(s_c)}
    \label{eq:ar_loss}
\end{equation}
where $s_c = \log p(o_c \mid x_j;\, \theta + \Delta\theta)$ is the log-likelihood of option $o_c$.

\paragraph{Next-Token Loss (NTL).}
The control condition applies standard causal language modeling:
\begin{equation}
    \mathcal{L}_\mathrm{NTL} = -\frac{1}{T} \sum_{t=1}^{T} \log p(x_t \mid x_{<t};\, \theta + \Delta\theta)
    \label{eq:ntl_loss}
\end{equation}
If only AR closes the gap, the recovery is specific to the belief-to-behavior mapping, supporting~\textbf{H2}.

\subsection{Implementation}
\label{sec:implementation}

At the start of each episode, we attach rank-$r$ LoRA adapters~\citep{hu2022lora} to the top 25\% of transformer layers and initialize them to zero.
Adaptation uses $K$ gradient steps with AdamW~\citep{loshchilov2018decoupled}; both LoRA weights and optimizer state are reset per episode.
We keep this configuration fixed across AR and NTL comparisons so that any recovery can be attributed to the supervision signal rather than benchmark-specific tuning.
Default hyperparameters: $r{=}8$, $\alpha{=}16$, $K{=}3$, $\eta{=}10^{-4}$.

\paragraph{Composition with prompting.}
Because BSTTT modifies weights and prompting modifies inputs, they compose naturally:
\begin{equation}
    \hat{a} = \argmax_{c} \;\; \log p\!\left(o_c \;\middle|\; f_\mathrm{prompt}(x_q);\; \theta + \Delta\theta_\mathrm{BSTTT}\right)
    \label{eq:composition}
\end{equation}
where $f_\mathrm{prompt}$ denotes any input transformation (CoT, SimToM, SymbolicToM, or Scratchpad).


%% ===== SECTION 4: EXPERIMENTAL SETUP =====
\section{Experimental Setup}
\label{sec:setup}

\subsection{Benchmarks}

\begin{table}[t]
    \caption{Benchmark characteristics. Bold question types = applied ($\mathcal{B}_\mathrm{A}$).}
    \label{tab:benchmarks}
    \centering
    \small
    \begin{tabular}{@{}lrccll@{}}
        \toprule
        \textbf{Benchmark} & $|\mathcal{B}|$ & $C$ & \textbf{Chance} & \textbf{Question types} & \textbf{Regime} \\
        \midrule
        SimpleToM & 60 & 2 & 50\% & Mental-state, \textbf{Behavior, Judgment} & Static, 1\textsuperscript{st}-order \\
        DynToM & 200 & 7--8 & $\sim$14\% & Understanding, \textbf{Transform-C, -D} & Temporal, multi-step \\
        Hi-ToM & 1{,}200 & 15 & 6.7\% & \textbf{Recursive belief, location} & Static, higher-order \\
        OpenToM & 2{,}000 & 3--6 & $\sim$25\% & Attitude, \textbf{Multihop first-order} & Static, multi-agent \\
        \bottomrule
    \end{tabular}
\end{table}

We evaluate across four benchmarks (Table~\ref{tab:benchmarks}) covering static/temporal scenarios, first-/higher-order belief, and recognition/application.

\subsection{Models}

We evaluate three instruction-tuned open-weight models: \textbf{Qwen-2.5-7B-Instruct}~\citep{qwen2.5}, \textbf{Llama-3.1-8B-Instruct}~\citep{grattafiori2024llama}, and \textbf{GPT-oss-20B}~\citep{openai2025gptoss120bgptoss20bmodel}, spanning three architectures and training data mixtures.

\subsection{Method Configurations}

\begin{table}[t]
    \caption{Method configurations and their diagnostic purpose.}
    \label{tab:methods}
    \centering
    \small
    \begin{tabular}{@{}lcccl@{}}
        \toprule
        \textbf{Method} & \textbf{Adapts $\theta$?} & \textbf{Prompt} & \textbf{Objective} & \textbf{Diagnostic purpose} \\
        \midrule
        \multicolumn{5}{@{}l}{\textit{Frozen baselines}} \\
        Frozen          & \ding{55} & Vanilla       & ---     & Baseline gap \\
        CoT             & \ding{55} & Step-by-step  & ---     & Reasoning elicitation \\[3pt]
        \multicolumn{5}{@{}l}{\textit{Structured prompting}} \\
        Scratchpad      & \ding{55} & Belief scaffold  & ---  & Externalized tracking \\
        SimToM          & \ding{55} & 2-stage perspective & --- & Perspective filtering \\
        SymbolicToM     & \ding{55} & Symbolic tracker   & --- & State representation \\[3pt]
        \multicolumn{5}{@{}l}{\textit{Diagnostic adaptation}} \\
        BSTTT-NTL       & \ding{51} & Vanilla       & NTL    & Control: generic exposure \\
        BSTTT-AR        & \ding{51} & Vanilla       & AR     & Treatment: ToM-aligned \\[3pt]
        \multicolumn{5}{@{}l}{\textit{Composed (adaptation + prompt)}} \\
        Scratchpad+BSTTT & \ding{51} & Scratchpad & AR  & Complementarity \\
        SimToM+BSTTT     & \ding{51} & SimToM     & AR  & Complementarity \\
        SymbolicToM+BSTTT & \ding{51} & SymbolicToM & AR & Complementarity \\
        \bottomrule
    \end{tabular}
\end{table}

We compare ten configurations spanning three categories (Table~\ref{tab:methods}).
All BSTTT variants use LoRA rank $r{=}8$, $\alpha{=}16$, top 25\% of layers, AdamW with $\eta{=}10^{-4}$, $K{=}3$ steps.
Seed coverage is benchmark-specific in the preserved artifact set used for this draft.
Unless otherwise noted, we report means over the preserved runs available for each model--benchmark--method cell; the core Hi-ToM and OpenToM comparisons use three recovered seeds, while SimpleToM and DynToM use the preserved multi-seed bundles available for those settings.


%% === INPUT PART 2 ===
\input{main_part2}

\end{document}
